SPAGHETTI Optimizes the Quantitative Analysis of \textit{In Vitro} Phase Contrast Miscroscopy through a Generative Deep Learning Model

### 2024-11-27 paper.tex
finish intro draft

Similar problems have been faced by pathologists in the analysis of histopathology images before. Hematoxylin and eosin (H\&E) staining is the most common staining method used in pathology to visualize the tissue structures and cell morphology and can provide a lot of rich information about the tissue disease state. However, for a long time, the information that can be extracted from H\&E images is limited to the human eye, and the analysis of H\&E images is often qualitative and subjective \autocite{he_stain}. Thanks to the development of deep learning, various approaches have been proposed to develop foundational feature extractors to extract the features from H\&E images and use them for downstream tasks such as making diagnosis and survival prediction. Recently, some attempts have even been made to make inferences of the gene expressions from H\&E images, suggesting a vast amount of information can be extracted by deep learning by looking at the cellular morphologies to make H\&E analysis more quantitative. \todo{add citations}

Order of the whole paper: Abstract, Introduction, Results, Discussion, Methods,
Data Availability, Code Availability, Acknowledgments, Funding, Author
Contributions, Competing Interests, Supplementary Information

There is where the intro
goes~\autocite{schwartzTooManyCellsIdentifiesVisualizes2020}. This includes a
short literature review highlighting the gaps that your paper fills in. Ends
with a summary of the paper.

>>>>

Similar problems have been faced by pathologists in the analysis of histopathology images before. Hematoxylin and eosin (H\&E) staining is the most common staining method used in pathology to visualize the tissue structures and cell morphology and can provide a lot of rich information about the tissue disease state. However, for a long time, the information that can be extracted from H\&E images is limited to the human eye, and the analysis of H\&E images is often qualitative and subjective \autocite{he_stain}. However recently, thanks to the development of deep learning, various approaches have been proposed to develop good performing foundational feature extractors \autocite{hoptimus0,phikonv2} to extract the features from H\&E images and use them for downstream tasks such as making diagnosis and survival prediction \autocite{survival_he_1,he_pdl1,he_disease_outcome}. Some attempts have even been made to make inferences of the gene expressions from H\&E images \autocite{bleep,hest-1k}, suggesting a vast amount of information can be extracted by deep learning by looking at the cellular morphologies to make H\&E analysis more quantitative.

Hence, the natural progression is to apply the same idea to PCM images by training a foundational feature extractor model for PCM. However, unlike H\&E images that can be found in rich resources such as the TCGA database \autocite{tcga}, the lack of publicly available PCM datasets makes this task challenging as a good performing foundational feature extractor model requires a large amount of data to train. To overcome this challenge, we hypothesize that instead of training a feature extractor for PCM from scratch, we can train a generative model to translate PCM images to H\&E-like images and use the foundational feature extractors that are already trained on H\&E images to extract the features from the translated images. Architectures such as cycle-consistent generative adversarial network (cycleGAN) \autocite{cyclegan2017} have been shown to be able to translate images from one domain to another without the need of paired images, and have been used in various tasks in the field of medical image translation \autocite{cyclegan_he,cyclegan_he_2}. We therefore developed \textit{SPAGHETTI (Ssim-restrained PhAse contrast Gan to H\&E TranslaTion of Images)}, a generative model that extends upon the original cycleGAN architecture to translate PCM images to H\&E-like images for further analysis. We showed that by using SPAGHETTI as a pre-processing step, we can improve the performance of the foundational H\&E feature extractors on PCM images on various quantitative downstream tasks such as classification and segmentation. SPAGHETTI is freely available at \url{https://github.com/schwartzlab-methods/spaghetti}.
